# Report (incomplete)

## Motivation
We want to analyse and compare Linear/Nonlinear and Spatial/Temporal/Spatio-Temporal relationships of windspeed data and electricity production from a nearby wind turbine.

## Data
We will use windspeed data from 6 different stations:

note: cmd+option+i
```{r}
print(data_explainer)
```

## New Features
From these data, we create lagged version of features for spatio-temporality. Also we create powered version of features for nonlinearity. etc..

In total we will have 4 different datasets. We will compare them to see the importance of temporal and nonlinear features. 

Also these datasets will be of use in comparing dimension reduction techniques.

1. Base Data

2. Lagged Data

3. Powered Data: ...

4. Lagged and Powered Data: ...

## Models
Here are the results. One can see that in all 6 stations the performance order of the data formats are same.

```{r}
results[order(station)]
```

lagged+powered >> powered >> lagged >> base

## Dimension Reduction Methods

We can see that adding new features increases the dimension very fast. We should do something about it.

1. Principal Component Analysis (PCA)

2. Kronecker PCA

3. Sparse Kronecker PCA
